---
title: "Premature mortality and substance use"
format:
  docx:
    reference-doc: ohid-template-november-2024.docx
editor: visual
---


- Ask 1: What proportion of premature deaths, given a cutoff point e.g. 55, are due to drugs and alcohol.
- Ask 2: Years of life lost (YLL) as a measure of premature mortality
- Ask 3: Add IMD


## 1. Proportion of deaths

### Number of deaths associated with drug use 

The ONS classifies death related to drug poisoning according ICD-10 codes. Certain ICD-10 codes classify a death as a "drug misuse death"[^1]. Each of these requires a specific substance (e.g. heroin) or substance category (e.g. opioids) to be indicated either in the ICD-10 code or on the death certificate. 

---

There are deaths each year where the ONS holds no information on the substance(s) involved:

| Year of death registration | All drug poisonings | Number of deaths without substance information | Percentage without substance information |
| :------------------------: | :-----------------: | :--------------------------------------------: | :---------------------------------------: |
|            2023            |       5,448         |                     1,245                      |                  22.9                     |
|            2022            |       4,907         |                     1,239                      |                  25.2                     |
|            2021            |       4,859         |                     1,219                      |                  25.1                     |
|            2020            |       4,561         |                     1,050                      |                  23.0                     |

---

Some of these will be classified as related to drug misuse where an ICD-10 code indicates mental and behavioural disorders due to drug use (excluding alcohol and tobacco) without a specific substance (e.g. F19 "multiple drug use and use of other psychoactive substances"). But others, broadly those coded as accidental/intentional self-poisonings or self-poisonings of unknown intent, will not be classfified as related to drug misuse unless a controlled drug under Misuse of Drugs Act 1971 was mentioned on the death record. 

The data linkage between ONS mortality and NDTMS allows some of those deaths to be identified indirectly as related to drug misuse where the person that died had had contact with the drug treatment system within a year of their date of death. 

#### National data, all ages

---

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=20, fig.height=18, fig.dpi=300}


library(flextable)

source("R/drug_deaths_functions.R")
source("R/plotting_functions.R")
source("R/themes.R")
source("R/dhsc_colour_palette.R")

drug_poisoning_deaths_file <-
  "data/raw/ndtms_mortality_data.parquet"
deaths_in_treatment_file <-
  "data/raw/tx_deaths_la_2122_2223.parquet"

national_data <- 
combine_national_data(
  poisoning_data = process_poisoning_data(file_path = drug_poisoning_deaths_file,
                                          date_of = "occurence",
                                          years = 2022),
  treatment_deaths_data = process_deaths_in_treatment(
    file_path = deaths_in_treatment_file,
    years = 2022,
    exclude_poisoning = TRUE,
    by_treatment_status = TRUE,
    by_death_cause = FALSE
  )
)

national_data <- 
  relabel_national_data(national_data = national_data)

p1 <- plot_national_data(plot_data = national_data)


p1 <- add_plot_annotations(plot = p1, data = national_data)

p1 + theme(text = element_text(size = 24))

```

---



```{r,echo=FALSE, message=FALSE, warning=FALSE}
ft <- 
 create_cause_of_death_table(
  process_deaths_in_treatment(
    file_path = deaths_in_treatment_file,
    years = 2022,
    exclude_poisoning = TRUE,
    by_treatment_status = TRUE,
    by_death_cause = TRUE
  )
) 
 
ft <- set_caption(ft, caption = "Causes of death")

ft

```
---

#### National data, by age group

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=20, fig.height=18, fig.dpi=300}

library(tidyverse)
library(afcolours)
library(openxlsx)
library(curl)

# Getting all cause deaths data, initially to see which age groups to use for
# drug/alcohol deaths; then to compare.

df <-
  read_csv("data/processed/drug_deaths_national.csv")


if (!file.exists("data/raw/ons_deaths_data_2023.xlsx")) {
  url <-
    "https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/birthsdeathsandmarriages/deaths/datasets/deathsregisteredinenglandandwalesseriesdrreferencetables/2023/annualdeathregistrations2023.xlsx"
  
  curl_download(url = url,
                destfile = "data/raw/ons_deaths_data_2023.xlsx")
  
}


ons_all_deaths_age <-
  read.xlsx(
    xlsxFile = "data/raw/ons_deaths_data_2023.xlsx",
    sheet = 6,
    sep.names = "_",
    startRow = 5
  ) |>
  janitor::clean_names()

ons_all_deaths_age <-
  ons_all_deaths_age |>
  as_tibble()

period_age_all_cause <-
  # Deaths from all causes by year and age only.
  ons_all_deaths_age |>
  filter(
    year_of_registration %in% c(2022, 2023),
    sex != "All people",
    age_years != "All ages",
    marital_status == "All marital statuses"
  ) |>
  group_by(year_of_registration, age_years) |>
  summarise(number_of_deaths = sum(number_of_deaths))

period_age_all_cause <-
  period_age_all_cause |>
  mutate(age_years = as.numeric(str_remove(age_years, "[^\\d]+"))) |>
  # Exclude ages outside the range of the drug-related deaths figures
  filter(age_years >= 18) |>
  filter(age_years <= 100)


period_age_drd <-
  # All deaths we've categorized as drug-related, by year and age
  df |>
  group_by(period, age) |>
  summarise(count = sum(count))

colnames(period_age_all_cause) <-
  c("period", "age", "deaths_all_causes")

colnames(period_age_drd) <-
  c("period", "age", "deaths_related_to_drug_misuse")


data <-
  left_join(period_age_all_cause, period_age_drd) |>
  mutate(deaths_related_to_drug_misuse = zoo::na.fill(deaths_related_to_drug_misuse, 0))

cut_age_groups <- 
function(x) {
  cut(
    x,
    breaks = c(17, 24, 34, 44, 54, 64, 74, 84, Inf),
    labels = c(
      "18-24",
      "25-34",
      "35-44",
      "45-54",
      "55-64",
      "65-74",
      "75-84",
      "85+"
    ),
    right = TRUE
  )
}


p1 <- 
data |> 
  filter(age < 55) |> 
  rowwise() |> 
  mutate(age_group = cut_age_groups(age)) |> 
  group_by(period, age_group) |> 
  summarise(across(where(is.numeric), sum)) |> 
  group_by(age_group) |> 
  summarise(across(where(is.numeric), mean)) |>  
  mutate(deaths_all_causes = deaths_all_causes - deaths_related_to_drug_misuse) |> 
  pivot_longer(cols = deaths_all_causes:deaths_related_to_drug_misuse) |> 
  mutate(name = str_replace(name, "all_causes", "all_other_causes")) |> 
  mutate(name = snakecase::to_sentence_case(name)) |> 
  ggplot(aes(x = age_group, y = value)) + 
  geom_col(aes(fill = name), position = "fill", width = 0.5, colour = "black", alpha = 0.8) + 
  theme_bw() + 
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = af_colours(n = 4)) + 
  theme(
    text = element_text(size = 24),
    legend.position = "bottom",
    legend.justification = "left"
  ) + 
  labs(fill = NULL,
       y = NULL,
       x = "Age group",
       title = "Deaths related to drug misuse as a proportion of all deaths, by age group",
       subtitle = "Average of 2022 and 2023 data"
       )


p1

```


```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=20, fig.height=18, fig.dpi=300}
p2 <- 
data |> 
  filter(age < 55) |> 
  rowwise() |> 
  mutate(age_group = cut_age_groups(age)) |> 
  group_by(period, age_group) |> 
  summarise(across(where(is.numeric), sum)) |> 
  group_by(age_group) |> 
  summarise(across(where(is.numeric), mean)) |>  
  mutate(deaths_all_causes = deaths_all_causes - deaths_related_to_drug_misuse) |> 
  pivot_longer(cols = deaths_all_causes:deaths_related_to_drug_misuse) |> 
  mutate(name = str_replace(name, "all_causes", "all_other_causes")) |> 
  mutate(name = snakecase::to_sentence_case(name)) |> 
  ggplot(aes(x = age_group, y = value)) + 
  geom_col(aes(fill = name), position = "identity", width = 0.5, colour = "black", alpha = 0.8) + 
  theme_bw() + 
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = af_colours(n = 4)) + 
  theme(
    text = element_text(size = 24),
    legend.position = "bottom",
    legend.justification = "left"
  ) + 
  labs(fill = NULL,
       y = NULL,
       x = "Age group",
       title = "Count of deaths related to drug misuse deaths by all other causes, by age group",
       subtitle = "Average of 2022 and 2023 data"
  )
p2
```
[^1]: The criteria for this classification are described in `Box 2` of the `Definition` tab of the latest relese of *Deaths related to drug poisoning, England and Wales* available [here](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/deaths/datasets/deathsrelatedtodrugpoisoningenglandandwalesreferencetable).
[^2]: This is the first four rows of `Table 11` in the same release.
---

## 2. Years of life lost due to substance use  

Years of life lost (YLL) is a measure of the impact of premature mortality, helpfully defined by Public Health England [here](https://fingertips.phe.org.uk/static-reports/health-profile-for-england/definitions-regional.html#years-of-life-lost-yll)

Chudasama et al. (2022) investigated five methods for estimating YLL[^3]. The first two methods are feasible with the available data for YLL from drug use and alcohol specific deaths. Only the drug-related YLL could be segmented by geographical estimates of deprivation.

All five methods are detailed in the supplementary PDF [here](https://ars.els-cdn.com/content/image/1-s2.0-S0895435622001639-mmc1.pdf).

### Mortality data
- Deaths related to drug misuse as defined by the ONS[^4].
- Additional drugs related to drugs misuse that the ONS had insufficient data to classify, but can be inferred from the data linkage of ONS data with NDTMS data by [Better Outcomes through Linked data (BOLD)](https://www.gov.uk/government/publications/ministry-of-justice-better-outcomes-through-linked-data-bold) [^5].
- Alcohol specific deaths

[^3]: Chudasama, Y.V., Khunti, K., Gillies, C.L., Dhalwani, N.N., Davies, M.J., Yates, T., & Zaccardi, F. (2022). Estimates of years of life lost depended on the method used: tutorial and comparative investigation. Journal of Clinical Epidemiology, 150, pp. 42-50. Available at: https://doi.org/10.1016/j.jclinepi.2022.06.012 [Accessed 6 Nov. 2024].
[^4]: *Not publicly available*
[^5]: *Not publicly available*
